Finiteness in Cubical Type Theory

### 2020-06-25 sections/search.tex
intro to exhaustibility

As it happens, this kind of search is a very common method of proof automation
in dependently-typed languages like Agda.
Proofs of statements like ``the following function is associative''
\begin{agdalisting}
  \ExecuteMetaData[agda/Snippets/Bool.tex]{and-def}
\end{agdalisting}
can be tedious: the associativity proof in particular would take \(2^3 = 8\)
cases.
This is unacceptable!
There are only finitely many cases to examine, after all, and we're
\emph{already} on a computer: why not automate it?
A proof that \(\AgdaDatatype{Bool}\) is finite can get us much of the way to a
library to do just that.

@@

So our task is to rewrite proof obligations like the one in
Equation~\ref{bool-assoc-auto-proof} into ones which can reduce completely. 
As it turns out, we have already described the type of proofs which can ``reduce
completely'': \emph{decidable} proofs.
If we have a decision procedure over some proposition \(P\) we can run that
decision during type checking, because the decision procedure itself is a proof
that the decision will terminate.
In code, we capture this idea with the following pair of functions:
\begin{multicols}{2}
  \begin{agdalisting}
    \ExecuteMetaData[agda/Snippets/Bool.tex]{is-true}
  \end{agdalisting} \columnbreak
  \begin{agdalisting}
    \ExecuteMetaData[agda/Snippets/Bool.tex]{from-true}
  \end{agdalisting}
\end{multicols}
The first is a function which derives a type from whether a decision is
successful or not.
This function is important because if we use the output of this type at any
point we will effectively force the unifier to run the decision computation.
The second takes---as an implicit argument---an inhabitant of the type generated
from the first, and uses it to prove that the decision can only be true, and the
extracts the resulting proof from that decision.
All in all, we can use it like this:
\begin{agdalisting}
  \ExecuteMetaData[agda/Snippets/Bool.tex]{extremely-obvious}
\end{agdalisting}
This technique will allow us to automatically compute any decidable predicate.
\section{Omniscience}
So we now know what is needed of us for proof automation: we need to take our
proofs and make them decidable.

@@

There is also a universal form of omniscience, which we call exhaustibility.
\begin{definition}[Exhaustibility]
  We say a type \(A\) is exhaustible if, for any decidable predicate \(P\) on
  \(A\), the universal quantification of the predicate is decidable.
  \begin{equation}
    \left( \Pi {(x : A)} , \mathbf{Dec}(P(x)) \right) \rightarrow \mathbf{Dec} \left( \Pi {(x : A)} , P(x) \right)
  \end{equation}
\end{definition}

>>>>

As it happens, this kind of search is a very common method of proof automation
in dependently-typed languages like Agda.
Proofs of statements like ``the following function is associative''
\begin{agdalisting}
  \ExecuteMetaData[agda/Snippets/Bool.tex]{and-def}
\end{agdalisting}
can be tedious: the associativity proof in particular would take \(2^3 = 8\)
cases.
This is unacceptable!
There are only finitely many cases to examine, after all, and we're
\emph{already} on a computer: why not automate it?
A proof that \(\AgdaDatatype{Bool}\) is finite can get us much of the way to a
library to do just that.
\todo[inline]{These examples so far are pretty focused on the bool associativity
  example.
  I'm not sure I can think of a good way to put countdown in instead: will we
  try switch?
  Or will we keep the bool for this short bit?
  }

@@

So our task is to rewrite proof obligations like the one in
Equation~\ref{bool-assoc-auto-proof} into ones which can reduce completely. 
As it turns out, we have already described the type of proofs which can ``reduce
completely'': \emph{decidable} proofs.
If we have a decision procedure over some proposition \(P\) we can run that
decision during type checking, because the decision procedure itself is a proof
that the decision will terminate.
In code, we capture this idea with the following pair of functions:
\begin{multicols}{2}
  \begin{agdalisting}
    \ExecuteMetaData[agda/Snippets/Bool.tex]{is-true}
  \end{agdalisting} \columnbreak
  \begin{agdalisting}
    \ExecuteMetaData[agda/Snippets/Bool.tex]{from-true}
  \end{agdalisting}
\end{multicols}
The first is a function which derives a type from whether a decision is
successful or not.
This function is important because if we use the output of this type at any
point we will effectively force the unifier to run the decision computation.
The second takes---as an implicit argument---an inhabitant of the type generated
from the first, and uses it to prove that the decision can only be true, and the
extracts the resulting proof from that decision.
All in all, we can use it like this:
\begin{agdalisting}
  \ExecuteMetaData[agda/Snippets/Bool.tex]{extremely-obvious}
\end{agdalisting}
This technique will allow us to automatically compute any decidable predicate.
\section{Omniscience}
So we now know what is needed of us for proof automation: we need to take our
proofs and make them decidable.
In particular, our strategy will be to take a proposition which is decidable in
the individual case (like
\((x\;\AgdaFunction{∧}\;y)\;\AgdaFunction{∧}\;z\;\AgdaFunction{≡}\;x\;\AgdaFunction{∧}\;(y\;\AgdaFunction{∧}\;z)\)
for a given \(x\), \(y\), and \(z\)) and make it decidable in the universal case.
Whether this is possible or not is a property of the \emph{domain} of a
decidable predicate, and is described formally by the notion of ``Exhaustibility''.
\begin{definition}[Exhaustibility]
  We say a type \(A\) is exhaustible if, for any decidable predicate \(P\) on
  \(A\), the universal quantification of the predicate is decidable.
  \begin{equation}
    \left( \Pi {(x : A)} , \mathbf{Dec}(P(x)) \right) \rightarrow \mathbf{Dec} \left( \Pi {(x : A)} , P(x) \right)
  \end{equation}
\end{definition}
Perhaps surprisingly, it is not only finite types which can are exhaustible.
Certain infinite types can be
exhaustible~\cite{escardoInfiniteSetsThat2007}.

@@

There is also a universal form of omniscience, which we call exhaustibility.
